---
title: "Student Budget Analysis"
author: "Salome Collante Ramirez, Andrés Ramírez, Ethan Marshall"
date: "2026-03-05"
output: pdf_document
---
# Overview

This notebook analyzes monthly rent payments reported by college students in South Florida using 2024 American Community Survey microdata (IPUMS USA). The goal is to understand what student housing budgets look like across Miami-Dade, Broward, and Palm Beach counties, and to classify students into three budget tiers (Cheap, Standard, and Luxury) based on their reported rent.

The resulting `price_category` label will serve as a join key connecting this dataset to a separate housing_data, enabling analysis of what housing options are realistically available at each budget level.

---

# Data

### About Data:

Loading Data
```{r loading data}
rm(list = ls())

# Working  Directory
setwd("~/Stats/Project - Housing/Final Drafts - Student Housing")

# Data
student_data = read.csv('Student_Budget.csv')

student_data$grade_level = ifelse(student_data$grade_level == 15, "Undergraduate", "Graduate")
```

Missing Values
```{r Missing Values}
colSums(is.na(student_data))
```

### Descriptive information:
```{r Descriptive Info}
names(student_data)
head(student_data)
summary(student_data)
dim(student_data)
```
Determining outliers

```{r}
# Boxplot with individual points
boxplot(student_data$rent,
        main = "Rent Distribution with Outliers",
        ylab = "Monthly Rent ($)",
        col  = "#2E75B6",
        border = "black",
        outline = FALSE,   # hides default outlier dots so we can color them manually
        ylim = c(0, 9000))   # extend to show the 8200 values

# All points
stripchart(student_data$rent, 
           vertical = TRUE, 
           method = "jitter",
           add = TRUE, 
           pch = 20, 
           col = rgb(0,0,0,0.2))

# Highlight the two extreme values in red
outlier_vals = student_data$rent[student_data$rent == 4 | student_data$rent == 8200]
stripchart(outlier_vals, 
           vertical = TRUE, 
           add = TRUE, 
           pch = 20, 
           col = "red",
           cex = 1.5)
```

```{r}
student_data = student_data[student_data$rent > 4 & student_data$rent < 5000, ]
cat("Rows remaining:", nrow(student_data))

attach(student_data)
```

Removed 11 observations with rent values below $10 or above $5,000, as these values are clearly separated from the rest of the distribution with no observations between $4,800 and $8,200.

### Descriptive Analysis:
Distribution of monthly rent across all students.
```{r rent distribution}
# Histogram- Rent Dist.
hist(rent,
     main = "Distribution of Monthly Rent - South Florida Students",
     xlab = "Monthly Rent ($)",
     col  = "#2E75B6",
     border = "white",
     breaks = 30,
     xlim = c(0, 5000))
abline(v =median(rent),col = "red",lwd = 2,lty=2)
legend("topright",legend=paste("Median: $",median(rent)),col= "red",lty = 2,lwd = 2)
```

Rent by county.
```{r rent by county}
# Box Plot- Rent by county
boxplot(rent ~ county,
        data = student_data,
        main = "Monthly Rent by County",
        xlab = "County",
        ylab = "Monthly Rent ($)",
        col  = c("#2E75B6", "#70AD47", "#ED7D31"))
```

Rent by grade level (Undergraduate | Graduate).
```{r rent by grade level}
# Boxplot- Rent by grade level
boxplot(rent ~ grade_level,
        data = student_data,
        main = "Monthly Rent by Grade Level",
        xlab = "Grade Level",
        ylab = "Monthly Rent ($)",
        col  = c("#2E75B6", "#ED7D31"))
```

Summary statistics by (1)county, (2)grade level.
```{r summary by county}
tapply(rent, county, summary)
tapply(rent, grade_level, summary)
```

### Budget Tier Assignment:

Students are classified into three budget tiers based on their monthly rent.

| Tier | Rent Range | Description |
|------|------------|-------------|
| Cheap | \$0 – \$1,600 | Shared housing, budget apartments |
| Standard | \$1,601 – \$2,000 | Mid-range apartments |
| Luxury | \$2,001+ | High-end or solo units |

```{r price category}
student_data$price_category = cut(
  student_data$rent,
  breaks = c(0, 1600, 2000, 5000),
  labels = c("Cheap", "Standard", "Luxury")
)

summary(student_data$price_category)
```

Distribution of budget tiers.
```{r tier distribution}
counts = table(student_data$price_category)
barplot(counts,
        main = "Student Budget Tier Distribution",
        xlab = "Budget Tier",
        ylab = "Number of Students",
        col  = c("#2E75B6", "#70AD47", "#ED7D31"),
        border = "white")
```

Tier breakdown by county.
```{r tier by county}
table(student_data$county,student_data$price_category)
```

Tier breakdown by grade level.
```{r tier by grade level}
table(student_data$grade_level,student_data$price_category)
```

### Data Preparation for Modeling:
```{r data preparation}
# Using age, and county as predictors
data_class = student_data[, c("rent","age","county","price_category")]
data_class$county =  as.factor(data_class$county)
data_class$price_category =  as.factor(data_class$price_category)

data_class = na.omit(data_class)

dim(data_class)
```

### Train / Test Split:
```{r train/test split}
set.seed(1)

# 80% of observations for training
num_samples = nrow(data_class)
train = sample(num_samples,num_samples * 0.8)

dim(data_class[train, ])
dim(data_class[-train, ])
```

```{r scale}
x = model.matrix(price_category ~.,data=data_class)[, -1]
train.x = x[train, ]
test.x = x[-train, ]

# Standardize variables (mean=0, sd=1)
train.x = scale(train.x)
test.x = scale(test.x,
               center = attr(train.x,"scaled:center"),
               scale  = attr(train.x,"scaled:scale"))

# Replace NaN from zero-variance columns with 0
train.x[is.nan(train.x)] = 0
test.x[is.nan(test.x)] = 0

train.y = data_class$price_category[train]
test.y = data_class$price_category[-train]
```

### 10-Fold Cross Validation:
```{r 10 fold cv, message=F}
library(class)
library(MASS)
library(randomForest)
library(pROC)

k = 10
set.seed(1)

# Vector that allocates each observation to one of k = 10 folds
folds = sample(1:k, nrow(data_class[train, ]), replace = TRUE)

# Matrix for results
cv.errors = matrix(NA,k,3,dimnames=list(NULL, c("LDA", "KNN", "RF")))


for (j in 1:k) {

  # KNN
  cv.train.x = train.x[folds != j, ]
  cv.test.x  = train.x[folds == j, ]
  cv.train.y = train.y[folds != j]
  cv.test.y  = train.y[folds == j]

  knn.cv = knn(cv.train.x,cv.test.x,cl=cv.train.y,k = 5)
  cv.errors[j, "KNN"]=mean(knn.cv==cv.test.y)

  # LDA
  cv.train.df = data_class[train, ][folds != j, ]
  cv.test.df  = data_class[train, ][folds == j, ]

  lda.cv = lda(price_category ~.,data=cv.train.df)
  lda.cv.pred = predict(lda.cv,cv.test.df)$class
  cv.errors[j,"LDA"] = mean(lda.cv.pred==cv.test.y)

  # Random Forest
  rf.cv = randomForest(price_category ~.,data=cv.train.df,ntree=500)
  rf.cv.pred = predict(rf.cv,newdata=cv.test.df)
  cv.errors[j,"RF"] = mean(rf.cv.pred==cv.test.y)
}

# Average over columns
mean.cv.errors = apply(cv.errors,2,mean)
mean.cv.errors
```

### KNN:
```{r knn}
# KNN = 1
knn.pred = knn(train.x, test.x, train.y, k=1)
train.knn.pred = knn(train.x, train.x, train.y, k=1)

# Confusion matrix
conf_matrix = table(knn.pred,test.y)
print(conf_matrix)

# Accuracy score
knn_train_acc = mean(train.knn.pred == train.y)
knn_test_acc = mean(knn.pred == test.y)

# AUC score
knn.pred.auc = knn(train.x, test.x, train.y, k=1, prob=TRUE)
knn.prob.auc = attr(knn.pred.auc, "prob")
knn.test.roc = multiclass.roc(test.y, knn.prob.auc)
knn_test_auc = auc(knn.test.roc)

cat('Train accuracy: ', knn_train_acc, "\n")
cat('Test accuracy: ', knn_test_acc, "\n")
cat('Test AUC: ', knn_test_auc, "\n")

# K = 11
knn.11.pred = knn(train.x, test.x, train.y, k=11)
train.11.pred = knn(train.x, train.x, train.y, k=11)

# Confusion matrix
conf_matrix = table(knn.11.pred,test.y)
print(conf_matrix)

# Accuracy score
knn11_train_acc = mean(train.11.pred == train.y)
knn11_test_acc = mean(knn.11.pred == test.y)

# AUC score
knn11.pred.auc = knn(train.x, test.x, train.y, k=11, prob=TRUE)
knn11.prob.auc = attr(knn11.pred.auc, "prob")
knn11.test.roc = multiclass.roc(test.y, knn11.prob.auc)
knn11_test_auc = auc(knn11.test.roc)

cat('Train k = 11 accuracy: ', knn11_train_acc, "\n")
cat('Test k = 11 accuracy: ', knn11_test_acc, "\n")
cat('Test k = 11 AUC: ', knn11_test_auc, "\n")
```

### LDA:
```{r lda}
lda.fit = lda(price_category ~.,data=data_class,subset=train)
lda.pred = predict(lda.fit, data_class[-train, ])
train.pred = predict(lda.fit, data_class[train, ])

# Confusion matrix
conf_matrix = table(lda.pred$class, test.y)
print(conf_matrix)

# Accuracy score
lda_train_acc = mean(train.pred$class == train.y)
lda_test_acc = mean(lda.pred$class == test.y)

# AUC score
lda_test_roc = multiclass.roc(test.y, lda.pred$posterior)
lda_test_auc = auc(lda_test_roc)

cat('Train accuracy: ', lda_train_acc, "\n")
cat('Test accuracy: ', lda_test_acc, "\n")
cat('Test AUC: ', lda_test_auc)
```

### Random Forest:
```{r random forest}
set.seed(1)
rf.fit = randomForest(price_category ~.,data=data_class,subset=train,ntree=500,importance=T)
yhat.rf = predict(rf.fit, newdata=data_class[-train, ])
train.rf = predict(rf.fit, newdata=data_class[train, ])

# Confusion matrix
conf_matrix = table(yhat.rf, test.y)
print(conf_matrix)

# Accuracy score
rf_train_acc = mean(train.rf == train.y)
rf_test_acc = mean(yhat.rf == test.y)

# AUC score
yhat.rf.auc = predict(rf.fit, newdata=data_class[-train, ], type='prob')
rf_test_roc = multiclass.roc(test.y, yhat.rf.auc)
rf_test_auc = auc(rf_test_roc)

cat('Train accuracy: ', rf_train_acc, "\n")
cat('Test accuracy: ', rf_test_acc, "\n")
cat('AUC score:', rf_test_auc)
```

Variable Importance
```{r variable importance}
importance(rf.fit)
varImpPlot(rf.fit)
```

### Model Comparison Summary:
```{r model comparison}
# Model names
models = c("LDA", "KNN_1", "KNN_11", "RF")

# Training accuracy
train_acc = c(lda_train_acc, knn_train_acc, knn11_train_acc, rf_train_acc)

# Test accuracy
test_acc = c(lda_test_acc, knn_test_acc, knn11_test_acc, rf_test_acc)

# Mean 10-fold CV 
fold10 = c(mean.cv.errors["LDA"], mean.cv.errors["KNN"], mean.cv.errors["KNN"], mean.cv.errors["RF"])

# AUC scores
auc_val = c(lda_test_auc, knn_test_auc, knn11_test_auc, rf_test_auc)

summary_table = data.frame(
  Model = models,
  Train_Acc = train_acc,
  Test_Acc = test_acc,
  Fold10 = fold10,
  AUC = auc_val
)

print(summary_table)
```

Our ROC results were:
```{r roc curves, warning=FALSE}
# Cheap vs all ROC
plot.roc(roc(test.y, lda.pred$posterior[,1]), col="blue", main="CHEAP vs all ROC")
lines.roc(roc(test.y, knn.test.roc$rocs[[1]]$original.predictor), col="red")
lines.roc(roc(test.y, knn11.test.roc$rocs[[1]]$original.predictor), col="yellow")
lines.roc(roc(test.y, yhat.rf.auc[,1]), col="green")
legend("bottomright", legend=c("LDA","KNN k=1", "KNN k=11","Random Forest"),
       col=c("blue","red","yellow","green"), lwd=2)

# Standard vs all ROC
plot.roc(roc(test.y, lda.pred$posterior[,2]), col="blue", main="Standard vs all ROC")
lines.roc(roc(test.y, knn.test.roc$rocs[[2]]$original.predictor), col="red")
lines.roc(roc(test.y, knn11.test.roc$rocs[[2]]$original.predictor), col="yellow")
lines.roc(roc(test.y, yhat.rf.auc[,2]), col="green")
legend("bottomright", legend=c("LDA","KNN k=1", "KNN k=11","Random Forest"),
       col=c("blue","red","yellow","green"), lwd=2)

# Luxury vs all ROC
plot.roc(roc(test.y, lda.pred$posterior[,3]), col="blue", main="Luxury vs all ROC")
lines.roc(roc(test.y, knn.test.roc$rocs[[3]]$original.predictor), col="red")
lines.roc(roc(test.y, knn11.test.roc$rocs[[3]]$original.predictor), col="yellow")
lines.roc(roc(test.y, yhat.rf.auc[,3]), col="green")
legend("bottomright", legend=c("LDA","KNN k=1", "KNN k=11","Random Forest"),
       col=c("blue","red","yellow","green"), lwd=2)
```
